---
title: "Beyond tidy data"
output: html_document
date: "2024-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Throughout this course we have been using tidy data: every dataset is a table organized
so that each column is a variable and each row is an observation. This design  is very 
convenient for data visualizations and data analysis. The format makes it easy to 
associate a column with an asethetic on a graph,  a predictor or response variable 
in a regression, or a column in a clustering or multivariate analysis.

There are at least two important ways many data sets differ from this tidy format: there 
may be multiple tables that are related to each other, known as relational data, or the 
data can be more irregular than a rectangular table. These formats can be reshaped into
rectangular tidy data, but at a cost of some repetition or creating missing values, both of which 
make the data larger to store and often harder to change or update.

In this lesson I will show you an example of each dataset, explain why the non-tidy formats are 
useful, and show you how to reshape the data for the tools we have studied in this course
which demand tidy (rectangular) data.

## Relational data

The classic example of relational data is a business with a list of products, 
a list of customers, and a combination of both describing sales of products to customers.
Another common example is a list of students, a list of courses, and a list of courses
taken by students (and their grades or other registration data).

In both of these cases, you can see that there are data sets that are independent of 
each other (products and customers; students and courses) and very important relationships 
between the two tables of data (products sold to customers, courses taken by students).

Here we will work with data on flights in and out of New York City, in a database.
There are five separate tables: airlines, airports, planes, flights, and weather.

```{r}
library(tidyverse, quietly = TRUE)
library(dbplyr)
library(nycflights13)
nyc <- nycflights13_sqlite()
airlines <- tbl(nyc, "airlines")
airports <- tbl(nyc, "airports")
airplanes <- tbl(nyc, "planes")
flights <- tbl(nyc, "flights")
```

Take a few minutes to familiarize yourself with each of these tables.
You'll notice that they are not like normal tables (or tibbles) in R. The number of rows
is unknown (reported as ??). This is because the data are not read into R; you 
simply have access to them to make queries (using dplyr, which is translated into SQL
for you.) This is convenient because there are 16 airlines, 3322 airplanes, 1458 airports, 
and 336,776 flights.

```{r}
count(airports)
count(flights)
```

You should notice that there are common variables between some pairs of tables:

* airplanes and flights have codes to identify each airplane (tailnum)
* airports and flights three letter codes for airports (faa, origin, dest)
* airlines and flights have two letter codes for the airline (carrier)

These allow the details of each airport, airline, and airplane to be connected
to data about each flights.

We can use these tables to find all the United Airlines ("UA") flights to Minneapolis ("MSP")
on an Embrarer 145 (EMB-145, EMB-145LR, EMB-145XR).

First, let's get the United Airlines flights to Minneapolis (operated by their subsiduary ExpressJet Airlines Inc.):

```{r}
flights_ev_msp <- flights |> filter(carrier == "EV", dest == "MSP")
```

We can get all the Embrarer airplanes:

```{r}
planes |> filter(substr(model, 1, 3) == "EMB")
```

Then we will use a join to get all the flights on one of these planes.

```{r}
inner_join(
  planes |> filter(substr(model, 1, 3) == "EMB"),
  flights_ev_msp,
  by = "tailnum"
)

```

## Irregular data


You can download data from online accounts from many services. For example, Spotify will prepare a zip file of the
music and podcasts you have listened to over the past year (or longer),

JSON
- e.g. Spotify data

```{r}
library(jsonlite)
# sp0 <- read_json("~/Dropbox/Spotify Account Data/StreamingHistory_music_0.json")
spotify0 <- fromJSON("~/Dropbox/Spotify Account Data/StreamingHistory_music_0.json", flatten = TRUE)
spotify1 <- fromJSON("~/Dropbox/Spotify Account Data/StreamingHistory_music_1.json", flatten = TRUE)
spotify2 <- fromJSON("~/Dropbox/Spotify Account Data/StreamingHistory_music_2.json", flatten = TRUE)
spotify <- bind_rows(spotify0, spotify1, spotify2)
spotify |> count(trackName) |> arrange(-n)
spotify |> group_by(trackName) |> summarize(time = sum(msPlayed)/1000/3600) |> arrange(-time)
spotify |> mutate(time = ymd_hm(endTime)) |>
  summarize(min_date = min(time), max_date = max(time))
spotify |> mutate(time = ymd_hm(endTime), jd = yday(time)) |>
  group_by(jd) |>
  summarize(duration = sum(msPlayed)/1000/86400) |>
  ggplot(aes(jd, duration)) + geom_line() + theme_bw()

```



