# R Review {.unnumbered}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(DT)
library(readxl)
funs <- read_excel(here("lessons/static", "functions.xlsx"), sheet = 1)
```

This page summarizes the most important R functions, datasets, and packages described in the course.
You can sort the following table by object (function or dataset), lesson, or package. You can also search the table by function name or description. There is a simple example of how to use most functions. Take a look at the help pages for more examples.

This table will be updated and expanded throughout the term.

I strongly encourage you to make your own notes about functions you learn to use. 

```{r echo=FALSE}
#| column: page
funs |> select(-Type) |>
  arrange(Lesson, Object) |>
  select(Object, Description, Example, Lesson, Package, everything()) |>
  datatable(rownames = FALSE,
            options = list(pageLength = 25))
```

Note that the `tidyverse` package includes the following packages: 

```{r results='asis', echo=FALSE}
cat(paste(tidyverse_packages() |> setdiff("tidyverse"), collapse=", "))
```


